### 2021-05-24 concurrency/docs/theory/theory.tex
minor edits
work on hash table writeup

@@

\[
  \einfer{Ghost-Refinement-Exchange}{
    \stacktwo{
      \widehat{M} \refines_{\rel} \widehat{N}
    }{
      \forall p,q.~ \rel(b\cdot p, q) \implies \exists q' .~ \rel(b' \cdot p, q') \land (a \cdot q \mapsto a' \cdot q')
    }
  }{
    \dboxed{b}^{\alpha \refines_{\rel} \gamma} * \dboxed{a}^\gamma
      \viewshift
    \dboxed{b'}^{\alpha \refines_{\rel} \gamma} * \dboxed{a'}^\gamma
  }
\]

@@

@@

\newcommand{\langloop}{\textbf{loop}}
\newcommand{\until}{\textbf{until}}
\newcommand{\langif}{\textbf{if}}
\newcommand{\langthen}{\textbf{then}}
\newcommand{\langelse}{\textbf{else}}

@@

\newcommand{\Inv}{Inv}
\newcommand{\inv}{inv}
\newcommand{\inlinehoare}[1]{{\color{blue}\left\{#1\right\}}}

Let's use a struct $\rwlock := \{\exc: \Pointer, \rc: \Pointer\}$. Our implementation will be straightforward:

\begin{align*}
  \AcquireExc(\rwlock) ~\eqdef~
      & \langloop~\until~\CAS(\rwlock.\exc,0,1);\\
      & \langloop~\until~(!\rwlock.\rc) = 0\\
      \\
  \AcquireShared(\rwlock) ~\eqdef~
      & \langloop~\until~(\\
        & ~~~~ (\langloop~\until~!\rwlock.\exc);\\
        & ~~~~ \AtomicAdd(\rwlock.\rc, 1);\\
        & ~~~~ \langif~ (!\rwlock.\exc) = 0 ~\langthen\\
        & ~~~~~~~~ \true\\
        & ~~~~ \langelse\\
        & ~~~~~~~~ \AtomicAdd(\rwlock.\rc, -1); \false\\
      )\\
      \\
  %\AcquireExcSetBit(\rwlock) ~\eqdef~
  %    & \CAS(\rwlock.\exc,\false,\true)\\
  %    \\
  %\AcquireExcCheckRc(\rwlock) ~\eqdef~
  %    & (!\rwlock.\rc) = 0\\
  %    \\
  %\AcquireSharedIncRc(\rwlock) ~\eqdef~
  %    & \AtomicAdd(\rwlock.\rc, 1)\\
  %    \\
  %\AcquireSharedAbort(\rwlock) ~\eqdef~
  %    & \AtomicAdd(\rwlock.\rc, -1)\\
  %    \\
  %\AcquireSharedCheckExc(\rwlock) ~\eqdef~
  %    & (!\rwlock.\exc) = \false\\
  %    \\
  \ReleaseExc(\rwlock) ~\eqdef~
      & \rwlock.\exc := 0\\
      \\
  \ReleaseShared(\rwlock) ~\eqdef~
      & \AtomicAdd(\rwlock.\rc, -1)
\end{align*}
We would like to prove the following specifications, for some predicate $\Inv(\rwlock,\alpha,\gamma,\inv)$. (Here, $\inv$ is the user-defined predicate on the ghost state stored by the $\rwlock$.)

@@

  \Inv(\rwlock) ~\eqdef~

@@

& \rwlock.\exc := 0\\

@@

>>>>

\newcommand{\BigAsterisk}{\mathop{\scalebox{3}{\raisebox{-0.2ex}{$\ast$}}}}%

@@

\[
  \einfer{Ghost-Refinement-Exchange}{
    \stacktwo{
      \widehat{M} \refines_{\rel} \widehat{N}
    }{
      \forall p,q.~ \rel(b\cdot p) = Some(q) \implies \exists q' .~ \rel(b' \cdot p) = Some(q') \land (a \cdot q \mapsto a' \cdot q')
    }
  }{
    \dboxed{b}^{\alpha \refines_{\rel} \gamma} * \dboxed{a}^\gamma
      \viewshift
    \dboxed{b'}^{\alpha \refines_{\rel} \gamma} * \dboxed{a'}^\gamma
  }
\]

@@

\newpage

@@

\newcommand{\langassign}{\leftarrow}
\newcommand{\langloop}{\textbf{loop}}
\newcommand{\until}{\textbf{until}}
\newcommand{\langif}{\textbf{if}}
\newcommand{\langthen}{\textbf{then}}
\newcommand{\langelse}{\textbf{else}}
\newcommand{\langmatch}{\textbf{match}}
\newcommand{\langwith}{\textbf{with}}
\newcommand{\langend}{\textbf{end}}
\newcommand{\langlet}{\textbf{let}~}
\newcommand{\langref}{\textbf{ref}}
\newcommand{\langin}{\textbf{in}}

@@

\newcommand{\Inv}{Inv}
\newcommand{\inv}{inv}
\newcommand{\HTInv}{HTInv}
\newcommand{\Heap}{Heap}
\newcommand{\htinv}{htinv}
\newcommand{\inlinehoare}[1]{{\color{blue}\left\{#1\right\}}}

Let's use a struct $\rwlock : \{\exc: \Pointer, \rc: \Pointer\}$. Our implementation will be straightforward:

\begin{align*}
  \AcquireExc(\rwlock) ~\eqdef~
      & \langloop~\until~\CAS(\rwlock.\exc,0,1);\\
      & \langloop~\until~(!\rwlock.\rc) = 0\\
      \\
  \AcquireShared(\rwlock) ~\eqdef~
      & \langloop~\until~(\\
        & ~~~~ (\langloop~\until~!\rwlock.\exc);\\
        & ~~~~ \AtomicAdd(\rwlock.\rc, 1);\\
        & ~~~~ \langif~ (!\rwlock.\exc) = 0 ~\langthen\\
        & ~~~~~~~~ \true\\
        & ~~~~ \langelse\\
        & ~~~~~~~~ \AtomicAdd(\rwlock.\rc, -1); \false\\
      )\\
      \\
  %\AcquireExcSetBit(\rwlock) ~\eqdef~
  %    & \CAS(\rwlock.\exc,\false,\true)\\
  %    \\
  %\AcquireExcCheckRc(\rwlock) ~\eqdef~
  %    & (!\rwlock.\rc) = 0\\
  %    \\
  %\AcquireSharedIncRc(\rwlock) ~\eqdef~
  %    & \AtomicAdd(\rwlock.\rc, 1)\\
  %    \\
  %\AcquireSharedAbort(\rwlock) ~\eqdef~
  %    & \AtomicAdd(\rwlock.\rc, -1)\\
  %    \\
  %\AcquireSharedCheckExc(\rwlock) ~\eqdef~
  %    & (!\rwlock.\exc) = \false\\
  %    \\
  \ReleaseExc(\rwlock) ~\eqdef~
      & \rwlock.\exc \langassign 0\\
      \\
  \ReleaseShared(\rwlock) ~\eqdef~
      & \AtomicAdd(\rwlock.\rc, -1)
\end{align*}
We would like to prove the following specifications, for some predicate $\Inv(\rwlock,\alpha,\gamma,\inv)$. (Here, $\inv$ is the user-defined predicate on the ghost state stored by the $\rwlock$.)

@@

  \Inv(\rwlock,\alpha,\gamma,\inv) ~\eqdef~

@@

& \rwlock.\exc \langassign 0\\

@@

\newpage

\textbf{A hash-table logic}

\newcommand{\hashto}{\stackrel{\text{hash}}{\hookrightarrow}}
\newcommand{\entryto}{\stackrel{\text{entry}}{\hookrightarrow}}
\newcommand{\Some}{\text{Some}}
\newcommand{\None}{\text{None}}

\newcommand{\Query}{\text{Query}}
\newcommand{\QueryIter}{\text{QueryIter}}
\newcommand{\Update}{\text{Update}}
\newcommand{\UpdateIter}{\text{UpdateIter}}
\newcommand{\Remove}{\text{Remove}}
\newcommand{\RemoveIter}{\text{RemoveIter}}
\newcommand{\NotInRange}{\text{NotInRange}}

\newcommand{\hash}{\ensuremath{hash}}
\newcommand{\rec}{\textbf{rec}}

%We will derive a TPCM (a ``normal'' one, not a refinement) to represent a hash table.
We'll want to support the following specs for a linear-probing hash table.

\begin{align*}
\inlinehoare{\HTInv(ht,\gamma) * \dboxed{k \hashto v_0}^\gamma} &~ \Query(ht,k) ~ \inlinehoare{v.~ \HTInv(ht,\gamma) * \dboxed{k \hashto v_0}^\gamma \land v = v_0}\\
\inlinehoare{\HTInv(ht,\gamma) * \dboxed{k \hashto v_0}^\gamma} &~ \Update(ht,k, v) ~ \inlinehoare{\HTInv(ht,\gamma) * \dboxed{k \hashto \Some~v}^\gamma}\\
\inlinehoare{\HTInv(ht,\gamma) * \dboxed{k \hashto v_0}^\gamma} &~ \Remove(ht,k) ~ \inlinehoare{\HTInv(ht,\gamma) * \dboxed{k \hashto \None}^\gamma}
\end{align*}

We assume a function $\hash : Keys \to \mathbb{N}$. We'll let $ht : \{rwlocks: rwlock~list, table: ((Key,Value)~option)~list\}$.

\begin{align*}
  \Query(ht,k) ~\eqdef~ & \QueryIter(ht,k,\hash(k))\\
  \QueryIter(ht,k,i) ~\eqdef~ & 
      \AcquireShared(ht.rwlocks[i]);\\
      &\langlet r = (\langmatch~ !ht.table[i] ~\langwith\\
      &~~~~ |~ \None \Rightarrow \None\\
      &~~~~ |~ \Some~(k_0,v_0) \Rightarrow
          \langif~ k = k_0 ~\langthen~ Some(v_0) ~\langelse~ \QueryIter(ht,k,i+1)\\
      &\langend) ~\langin\\
      &\ReleaseShared(ht.rwlocks[i]);\\
      &r\\
      \\
  \Update(ht,k,v) ~\eqdef~ & \UpdateIter(ht,k,v,\hash(k))\\
  \UpdateIter(ht,k,v,i) ~\eqdef~ & 
      \AcquireExc(ht.rwlocks[i]);\\
      &\langlet r = (\langmatch~ !ht.table[i] ~\langwith\\
      &~~~~ |~ \None \Rightarrow ht.table[i] \langassign (k,v)\\
      &~~~~ |~ \Some~(k_0,v_0) \Rightarrow
          \langif~ k = k_0 ~\langthen~ ht.table[i] \langassign (k,v) ~\langelse~ \UpdateIter(ht,k,v,i+1)\\
      &\langend) ~\langin\\
      &\ReleaseExc(ht.rwlocks[i]);\\
      &r
\end{align*}

\newpage

We'll define a TPCM resource for our hash table (that is, a ``normal'' TPCM, not a refinement). It will have two kinds of elements, $k \hashto v$ (where $v : Value~option$)
and $i \entryto c$, where $c : (Key,Value)~option$.

Allowable operations will be encoded into this logic. For example, we will have all the following:

\emph{Update}.

\begin{align*}
  &\hash(k) \le u ~\land~ \left(\forall i.~ \hash(k) \le i < u \implies i \in entries \land \exists k_1,v_1 .~ entries[i] = \Some (k_1,v_1) \land k_1 \ne k\right)\\
  &~~~~\land~ \left(entries[u] = \None ~\lor~ \exists v_1 .~ entries[u] = \Some (k, v) \right)\\
  &~~~~\implies entries \cdot (k \hashto v_0) ~\mapsto~ entries[u \langassign \Some~(k,v)]  \cdot (k \hashto v)
\end{align*}

\emph{Query (found)}.

The found case is easy: a single entry $(k,v)$ proves that the hash table maps $k$ to $v$.

\[
  \left((i \entryto (k,v_0)) \cdot (k \hashto v)\right) \ne \bot \implies v = \Some(v_0)
\]

The not-found case requires another quantifed predicate:

\begin{align*}
&\left(\forall i.~ \hash(k) \le i < u \implies i \in entries \land \exists k_1,v_1 .~ entries[i] = \Some (k_1,v_1) \land k_1 \ne k\right)\\
  &~~~~\land~ \left(entries[u] = \None\right)\\
  &~~~~\implies entries \cdot (k \hashto v) \ne \bot \implies v = \None
\end{align*}

We'll define $\HTInv$ as follows:
\begin{align*}
\HTInv(ht, \gamma) ~\eqdef~& \exists\alpha.~ \BigAsterisk_i \Inv\left(ht.rwlocks[i], \alpha, \gamma\times\Heap, \htinv(i, ht.entries[i])\right)\\
\htinv(i, \ell) ~\eqdef~& \lambda x .~ \exists c .~ x = (i \entryto c, \ell \pointsto c)
\end{align*}
In other words, the RwLocks will be arranged so that we can use the lock at index $i$ to obtain
some ghost state $(i \entryto c, ht.entries[i] \pointsto c)$.

\newpage

Now, we prove $\QueryIter$. (To avoid clutter, we silently skolemize existential
variables as they are introduced.)

\begin{align*}
  &\inlinehoare{\active(\kappa) * \borrow{\kappa}{\dboxed{ht.entries}^\gamma}
      * \dboxed{k \hashto v_0}^\gamma \land KeyNotInRange(ht.entries, k, i)}\\
  &\AcquireShared(ht.rwlocks[i]);\\
  &\inlinehoare{\active(\kappa) * \borrow{\kappa}{\dboxed{ht.entries}^\gamma}
      * \dboxed{k \hashto v_0}^\gamma \land KeyNotInRange(ht.entries, k, i)
      * \SharedGuard\left( (i \entryto c, ht.entries[i] \pointsto c) \right) }\\
  &\langlet r = (\langmatch~ !ht.table[i] ~\langwith\\
  &~~~~ |~ \None \Rightarrow \None\\
  &~~~~ |~ \Some~(k_0,v_0) \Rightarrow
      \langif~ k = k_0 ~\langthen~ Some(v_0) ~\langelse~ \QueryIter(ht,k,i+1)\\
  &\langend) ~\langin\\
  &\ReleaseShared(ht.rwlocks[i]);\\
  &r\\
  &\inlinehoare{\active(\kappa) * (k \hashto v_0) \land v = v_0}
\end{align*}
